---
title: "Does densily populated is New Zealand?"
author: "Adam J Campbell"
date: "12 May, 2020"
output: html_document
---

Load in USA data and libraries
```{r Initial_Loading, warning=FALSE, message=FALSE}

USA_2018_AREAS <- read.csv('data/tracts_2018.csv')
USA_2018_POP <- read.csv('data/nhgis0002_ds239_20185_2018_tract.csv')

library("tidyverse")
library("LaplacesDemon")
```

Join the population data to the area data.  At this point, we'll filter out the records with zero area and only select a few important columns.  We'll also change the area to km<sup>2</sup>.

```{r create_population_area}
population_area <- 
  right_join(USA_2018_POP, USA_2018_AREAS, by = "GISJOIN") %>%
  filter(ALAND > 0) %>%   # filter out areas that have zero land area
  mutate(LAND_AREA_km2 = ALAND / 1000000) %>%
  select(c("STATE", "TRACTA", "AJWME001", "LAND_AREA_km2"))
#population_area$LAND_AREA_km2 <- population_area$ALAND / 1000000
```

Now we need to load and process the New Zealand data into the format for `population_area` and append that data.
```{r}
NZ_data <-
  read.csv("data/NZ_export.csv") %>%
  filter(LAND_AREA_ > 0) %>% # filter out areas that have zero land area
  mutate(STATE = "New Zealand") %>% # add a state 
  rename(
    TRACTA = SA22018_V1,
    LAND_AREA_km2 = LAND_AREA_,
    AJWME001 = Pop_Tota_2
    ) %>%
  select(c("STATE", "TRACTA", "AJWME001", "LAND_AREA_km2"))
  
population_area <-rbind(population_area, NZ_data)

```


Create `state_stats` which are broad summary statistics for each state (plus a few other places) including total population and total area.

```{r create_state_stats}
state_stats <-
  population_area %>%
  group_by(STATE) %>%
  summarise(state_pop = sum(AJWME001), state_land_area = sum(LAND_AREA_km2)) %>%
  ungroup()

state_stats$population_density_land <- state_stats$state_pop / state_stats$state_land_area
```

Join the `state_stats` data onto the `population_area` data and calculate the fraction of population and land area each census tract occupies relative to its state.

```{r join_population_area}
population_area <-
  left_join(population_area, state_stats, by = "STATE")

population_area$land_area_frac <- population_area$LAND_AREA_km2 / population_area$state_land_area

population_area$pop_frac <- population_area$AJWME001 / population_area$state_pop
```

Now we will compute the KLD sum for each state.  

```{r message=FALSE, warning=FALSE}

KLD_stats<-
  population_area %>%
  group_by(STATE) %>%
  select(pop_frac, land_area_frac) %>%
  summarise(KLD = KLD(pop_frac, land_area_frac)$sum.KLD.px.py)

```

Join `KLD_stats` to `state_stats` and compute a few fields. 

```{r}
state_stats <- inner_join(state_stats, KLD_stats, by="STATE")

state_stats$log_density <- log(state_stats$population_density_land)
state_stats$density_KLD <- state_stats$log_density + state_stats$KLD
state_stats$lived_density <- exp(state_stats$density_KLD)
```

Make a new dataframe `state_density` that cleans up `state_stats` and keeps only items for display

```{r}
state_density <- 
  state_stats %>%
  select(STATE, state_pop, state_land_area, population_density_land, lived_density) %>%
  mutate(density_ratio = lived_density/population_density_land)
```


Save data frame
```{r}
saveRDS(state_density, file="state_density.Rda")
```




